---
title: "TP_NN"
author: "Loic Verlingue"
date: "13 janvier 2020"
output: html_document
---
# What you will do in this R Notebook

## Introduction
This is the second day of your training in R for Machine Learning. You will now enter the wide, wild and exciting field of Deep Learning!

Neural Networks and Deep Learning have revolutionized the performance of many Machine Learning task: in image analysis, gaming and natural langage understanding, fro example.

Again your input is a huge amount of variables (genes expression per patients) and you want to provide the clinician a simple output that he can use for his patients.

I hope you'll enjoye this Notebook and that it can help you to adress your current and/or future challenges.

## Main steps
After loading TCGA data, we will build and train a simple neural network.

## In practice
The code is provided to you. You will just have to follow the instructions all along the Notebook.
I'll give you a little bit of explanation when needed. 

Pay attention: there will be some questions to answer in this notebook.

Finally you will have a partial code to complete at the end of the notebook, and give it back to us.

*--> lets' go!*

## Functions
```{r}
library(keras)
set.seed(1234)
```

```{r}
#source("https://raw.githubusercontent.com/DITEP/TPBigDataIFSBM/master/LoadcBioportal150119.R")
source("https://raw.githubusercontent.com/gustaveroussy/IFSBM-bigdata/master/TP_IFSBM_module12/LoadcBioportal150119.R")
# install cgdsr
install.packages("cgdsr")
install.packages("devtools")
# download https://www.anaconda.com/distribution/#download-section
devtools::install_github("rstudio/keras")
library(keras)
install_keras()
# help for R keras at https://keras.rstudio.com/
library(reshape2)
```


## Data

Select the genes you want to use.
```{r}
urlfile<-"https://raw.githubusercontent.com/gustaveroussy/IFSBM-bigdata/master/TP_IFSBM_module12/CancerGeneCensusCOSMIC.csv"
CGS<-read.csv(urlfile,stringsAsFactors = F)
GENES<-CGS$Gene.Symbol[CGS$Hallmark=="Yes"]
print(paste("You have selected",length(GENES),"important cancer genes like:", paste(sample(GENES,20), collapse = ", "), "..."))
```

```{r}
TCGAdata<-LoadcBioportal(Genes = GENES, Organ = "luad|lusc", 
                         ClinicNeeded = T, MutNeeded = F, 
                         RNANeeded = T, NormalizeRNA = T, 
                         FunctionalAnnot = F, PDF = F, Tests=T)
STUDY<-TCGAdata$STUDY
print("Here are your patients' numbers:")
print(table(TCGAdata$CLINIC$study))
```

```{r}
dim(TCGAdata$EXP)
```
```{r}
TCGAdata$CLINIC$study_num<-as.numeric(as.factor(TCGAdata$CLINIC$study))-1
```


## Split the data
```{r}
TrainSplit=0.8
Train<-sample(seq(nrow(TCGAdata$CLINIC)),size = round(nrow(TCGAdata$CLINIC)*TrainSplit) )
Val<-seq(nrow(TCGAdata$CLINIC))[!seq(nrow(TCGAdata$CLINIC))%in%Train]
paste("Your training and validation cohorts have ",length(Train),"&", length(Val), "patients.")
```

## Model building
Building the neural network requires configuring the layers of the model, then compiling the model.

### Setup the layers
The basic building block of a neural network is the layer. Layers extract representations from the data fed into them. And, hopefully, these representations are more meaningful for the problem at hand.
Most of deep learning consists of chaining together simple layers. Most layers, like layer_dense, have parameters that are learned during training.
```{r}
model <- keras_model_sequential()
model %>%
  layer_flatten(input_shape = c(250)) %>%
  layer_dense(units = 128, activation = 'relu') %>%
  layer_dense(units = 1, activation = 'sigmoid')
```

### print your model architecture
```{r}
model
```


### Compile the model
Before the model is ready for training, it needs a few more settings. These are added during the model’s compile step:

Loss function — This measures how accurate the model is during training. We want to minimize this function to “steer” the model in the right direction.
Optimizer — This is how the model is updated based on the data it sees and its loss function.
Metrics —Used to monitor the training and testing steps. The following example uses accuracy, the fraction of the images that are correctly classified.
```{r}
model %>% compile(
  optimizer = 'adam', 
  loss = 'mean_squared_error',
  metrics = c('accuracy')
)
```


## Model training
Training the neural network model requires the following steps:

Feed the training data to the model — in this example, the train_images and train_labels arrays.
The model learns to associate images and labels.
We ask the model to make predictions about a test set — in this example, the test_images array. We verify that the predictions match the labels from the test_labels array.
To start training, call the fit method — the model is “fit” to the training data:

```{r}
model %>% fit(as.matrix(TCGAdata$EXP[Train,]),TCGAdata$CLINIC$study_num[Train], epochs = 100, validation_split = 0.2, shuffle = FALSE)
```

As the model trains, the loss and accuracy metrics are displayed. This model reaches an accuracy of about XX on the training data.

## Evaluate accuracy
Next, compare how the model performs on the test dataset:
```{r}
score <- model %>% evaluate(as.matrix(TCGAdata$EXP[Val,]), TCGAdata$CLINIC$study_num[Val])
cat('Test loss:', score$loss, "\n")
cat('Test accuracy:', score$acc, "\n")
```

## Make predictions
With the model trained, we can use it to make predictions about some new data.
```{r}
predictions <- model %>% predict(as.matrix(TCGAdata$EXP[Val,]))
```

Here, the model has predicted the label for each image in the testing set. Let’s take a look at the first prediction:

```{r}
predictions[1, ]
```


Alternatively, we can also directly get the class prediction:
```{r}
class_pred <- model %>% predict_classes(as.matrix(TCGAdata$EXP[Val,]))

```

```{r}
table(class_pred,TCGAdata$CLINIC$study[Val])
```
